---
title: Simulating synthetic data of same form as Milan COVID-19 analysis
---

From the paper by [Jackson et al.](https://bmjopen.bmj.com/content/12/3/e054859.abstract).

The original data are under strict restrictions, so here we simulate data of the same structure to use for work on statistical methodology. 


# Original data structure 

* 27598 people testing positive for SARS-CoV-2 between Feb-Jul 2020 and followed up to Jul 2020 

* Time to either hospital admission, death without hospital admission, or confirmed recovery are recorded.    For those admitted to hospital, dates of ICU admission, death or discharge from hospital are also recorded. 


## Predictors 

* Age (originally recorded as continuous, but treated as categorical in the analysis).  Median 64, interquartile range 48 - 82. 

* Sex (45% male)

* Occupational risk status (3 levels): healthcare workers (13%), care home residents (21%), and all others (66%)

* Presence of any comorbidities (only treated as binary in the analysis) (47%)

* Calendar date of onset (continuous, but treated as categorical in the analysis):  February, March, April, May or June.   1588, 14297, 8286, 2590, 837 people respectively.




# Models

Multinomial regression was used to describe the probability $p_r(x)$ that each competing event $r$ (hospital admission, death or recovery) happens first, as a function of covariates $x$. 

The time $T_r$ to the event $r$, given that this is the event that happens first, is modelled with a generalised gamma distribution.   The expected time is related to covariates $x$ with an accelerated failure time model. 

Selected covariates $x$ for each event and time to event, and their coefficients, are listed in Tables 13-17 of the appendix to the paper. 



# Simulation of synthetic data from the fitted models

## Times to events for baseline covariate category 

Median, lower and upper 95\% quantiles for days to admission, death and recovery, taken from the paper:

```{r}
library(tidyverse)
library(flexsurv)
dat <- tibble(
  lower = c(0.5, 0.7, 17),
  median = c(5.5, 9.3, 57),
  upper = c(38, 48, 131),
  event = c("adm","death","rec")
)
```

Estimate the parameters of a generalised Gamma distribution by a least squares fit to these quantiles.   These parameters refer to the baseline covariate category (age under 45, onset in March, no comorbidities, neither healthcare worker nor care home resident)

```{r}
p <- c(0.025, 0.5, 0.975)
ggfit <- function(event){
  obs <- unlist(dat[dat$event==event,1:3])
  sumsq <- function(par) sum((qgengamma(p, par[1], exp(par[2]), par[3]) - obs)^2)
  ggpars <- optim(c(0,0,0), sumsq)$par
  list(mu=ggpars[1], sigma=exp(ggpars[2]), Q=ggpars[3])
}
pars_adm <- ggfit("adm")
pars_death <- ggfit("death")
pars_rec <- ggfit("rec")

exec(rgengamma, n=10, !!!pars_adm)
exec(rgengamma, n=10, !!!pars_death)
exec(rgengamma, n=10, !!!pars_rec)
```


## Generate population with covariates with roughly same distributions as in paper

enerate age distribution from a gamma distribution with the same median as in the paper.  Make the quantiles narrower than the IQR presented in the paper, to ensure not too many people are over 100. 

```{r}
set.seed(1)
p <- c(0.1, 0.5, 0.9)
obs <- c(48, 64, 82)
sumsq <- function(par) sum((qgamma(p, exp(par[1]), exp(par[2])) - obs)^2)
gpars <- optim(c(0,0), sumsq)$par
pars_age <- list(shape=exp(gpars[1]), rate=exp(gpars[2]))
```

Generate COVID-19 onset days uniformly within months, given proportion of onsets in each month


```{r}
n <- 27598
monset <- c(1588,14297,8286,2590,837)
days <- c(feb=28, mar=31, apr=30, may=31, jun=30)
days_sim <- sample(1:sum(days), size=n, replace=TRUE, prob=rep(monset, days)) # 1 = feb 1st 
months_sim <- factor(cut(dat_unif, c(0, cumsum(days)), include.lowest = TRUE), labels=names(days))
```

Generate sex and comorbidity status independently of all other variables, given the proportions in the paper. 


```{r}
simdat <- tibble(
  age = exec(rgamma, n=n, !!!pars_age),
  sex = factor(sample(c("male","female"), n, replace=TRUE, prob=c(45, 55))),
  comorb = factor(sample(c("no","yes"), n, replace=TRUE, prob=c(53, 47))),
  day = days_sim,
  month = relevel(months_sim, "mar")
) |>
  mutate(agegroup = cut(age, c(0,45,65,Inf), labels = c("ageu45","age46","age66")))
```

Generate occupational status given the proportions presented in the paper, assuming care home residents are all over 65, and healthcare workers are all aged 65 or under. 

```{r}
occs <- c("neither","hcw","chr")
occ_tab <- round(n*c(66,13,21)/100) |> setNames(occs)
occ <- factor(sample(occs, n, replace=TRUE, prob=c(66,13,21)), levels=occs)
age66_ind <- which(simdat$agegroup=="age66")
chr_inds <- sample(age66_ind, occ_tab["chr"])
agerest_ind <- which(simdat$agegroup != "age66")
hcw_inds <- sample(agerest_ind, occ_tab["hcw"])
simdat$occ <- factor("neither", levels = occs)
simdat$occ[chr_inds] <- "chr"
simdat$occ[hcw_inds] <- "hcw"
head(simdat)
```




## Determine probability of each outcome for each individual 

To do this, we apply covariate effects to the baseline log odds of events.

Covariate effects are taken from Table 13 of the paper appendix, to produce a probability of each outcome for each individual.

Then sample an event with these probabilities

```{r}
padm_base <- 0.337 
pdeath_base <- 0.076
prec_base <- 1 - padm_base - pdeath_base
loadm_base <- log(padm_base / prec_base)
lodeath_base <- log(pdeath_base / prec_base)
coefs <- read.csv("app_tab13_tidy.csv")
```

The multinomial logistic regression models used for the two competing outcomes (time to admission, time to death) are then defined as R formula objects.

The "design matrix" for these models is obtained by applying the `model.frame` and  `model.matrix` functions to the simulated data `simdat`.

The linear predictors for the models can then be defined as $X\beta$, where $X$ is the model matrix, and $\beta$ are the coefficients of the regression model.

```{r}
form_adm <- ~ month + agegroup + sex + occ + comorb + month :agegroup + agegroup:sex + agegroup:occ + month:sex + month:occ
mf <- model.frame(form_adm,data=simdat)
Xadm <- model.matrix(form_adm, mf)
coef_adm <- coefs |> filter(event=="padm") |> select(cov,est) |> 
  bind_rows(list(cov="(Intercept)",est=loadm_base)) |>
  slice(match(colnames(Xadm), cov))
stopifnot(all(coef_adm$cov == colnames(Xadm)))
oddsadm <- exp(Xadm %*% coef_adm$est)

form_death <- ~ month + agegroup + sex + occ + comorb + agegroup:occ + month:occ
mf <- model.frame(form_death,data=simdat)
Xdeath <- model.matrix(form_death, mf)
coef_death <- coefs |> filter(event=="pdeath") |> select(cov,est) |> 
  bind_rows(list(cov="(Intercept)",est=lodeath_base)) |>
  slice(match(colnames(Xdeath), cov))
stopifnot(all(coef_death$cov == colnames(Xdeath)))
oddsdeath <- exp(Xdeath %*% coef_death$est)
```

This gives the odds of admission relative to recovery (odds(admission) = P(admission) / P(recovery)), and the odds of death relative to recovery (odds(death) = P(death) / P(recovery)).

The probability of each event can be determined, given that the probabilities of recovery, admission and death add up to 1.

```{r}
prec <- 1 / (1 + oddsadm + oddsdeath)
padm <- oddsadm*prec
pdeath <- oddsdeath*prec
stopifnot(all(prec <= 1 & prec >= 0))
stopifnot(all(padm <= 1 & padm >= 0))
stopifnot(all(pdeath <= 1 & pdeath >= 0))
stopifnot(isTRUE(all.equal(as.numeric(prec + padm + pdeath), rep(1, n))))
pevent <- cbind(recovery=prec, admission=padm, death=pdeath)
```

A random event is then simulated for each person in the data, with these probabilities.
```{r}
events <- c("recovery","admission","death")
res <- numeric(n)
for (i in 1:n) # this is actually fast!
  res[i] <- sample(events, prob=pevent[i,], size=1, replace=TRUE)

simdat$event_true <- factor(res, levels=events) 
```


## Sample times to events 

From generalized gamma models in the paper. 

Apply covariate effects on these first.

Apply censoring at 30 June (150 days after 1 Feb).

```{r}
form_adm <- ~ agegroup + sex + month + occ + comorb
mf <- model.frame(form_adm,data=simdat)
Xadm <- model.matrix(form_adm, mf)
coef_adm <- coefs |> filter(event=="tadm") |> select(cov,est) |> 
  bind_rows(list(cov="(Intercept)",est=pars_adm$mu)) |>
  slice(match(colnames(Xadm), cov))
stopifnot(all(coef_adm$cov == colnames(Xadm)))
mu_adm <- Xadm %*% coef_adm$est

form_death <- ~ agegroup + sex + month + occ + comorb
mf <- model.frame(form_death,data=simdat)
Xdeath <- model.matrix(form_death, mf)
coef_death <- coefs |> filter(event=="tdeath") |> select(cov,est) |> 
  bind_rows(list(cov="(Intercept)",est=pars_death$mu)) |>
  slice(match(colnames(Xdeath), cov))
stopifnot(all(coef_death$cov == colnames(Xdeath)))
mu_death <- Xdeath %*% coef_death$est

form_rec <- ~ agegroup + sex + month + occ + comorb + month:agegroup + occ:agegroup + month:occ
mf <- model.frame(form_rec,data=simdat)
Xrec <- model.matrix(form_rec, mf)
coef_rec <- coefs |> filter(event=="trec") |> select(cov,est) |> 
  bind_rows(list(cov="(Intercept)",est=pars_rec$mu)) |>
  slice(match(colnames(Xrec), cov))
stopifnot(all(coef_rec$cov == colnames(Xrec)))
mu_rec <- Xrec %*% coef_rec$est

tadm <- rgengamma(n, mu_adm, pars_adm$sigma, pars_adm$Q)
tdeath <- rgengamma(n, mu_death, pars_death$sigma, pars_death$Q)
trec <- rgengamma(n, mu_rec, pars_rec$sigma, pars_rec$Q)

simdat <- simdat |>
  mutate(time_true = if_else(event_true=="death", tdeath, ifelse(event_true=="admission", tadm, trec)),
         cens = if_else(day + time_true <= 150, "obs", "cens"),
         event = if_else(cens=="cens", NA, event_true), # ifelse removes factor 
         time = if_else(cens=="cens", 150, time_true))

saveRDS(simdat, file="data-raw/milansim.rds")

summary(simdat)

```
